Modeling mental rotation

### 2014-01-29 man/cogsci-2014/cogsci-2014.tex
More work on modeling section

\section{Models of mental rotation}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta,
  \label{eq:posterior}
\end{equation}
where $p(X_b\ \vert\ X_a, \theta, h)$ is the probability that $X_b$
was generated from $X_a$, and $p(X_a)$ is the probability of observing
$X_a$. Because we ultimately want to determine which hypothesis is
more likely, the quantity of interest is a ratio $\ell:=p(h=0\ \vert\
X_a, X_b) / p(h=1\ \vert\ X_a, X_b)$ which (assuming that all
rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\left(\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta\right)p(h=0)}{\left(\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta\right)p(h=1)}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

We represent a shape of $N$ vertices with a $N\times 2$ coordinate
matrix $\mathbf{X}=[\mathbf{x}_1, \ldots{}, \mathbf{x}_N]$, and denote
the rotation and/or reflection transformation as $f(\mathbf{X}, h,
\theta):=\mathbf{X}\F_h^T\R^T$, where $\R$ is a rotation matrix, and
$\F_h$ is either the identity matrix $\I$ (when $h=0$) or a reflection
matrix across the $y$-axis (when $h=1$).  We assume that the observed
shapes may only be transformed by a small amount at a time, and each
transformation is costly in terms of computational resources. The
goal, then is to estimate the integrals in Eq.~\ref{eq:lh-ratio} by
evaluating $p(\Xb\ \vert\ \Xa, \theta, h)$ as few times as possible.

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$:
\begin{equation}
  p(\Xb\ \vert\ \Xa, \theta, h):= S(\Xb, f(\Xa, h, \theta)).
  \label{eq:likelihood}
\end{equation}
We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings,
$\mathbb{M}$. For brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h,
\theta)$ where $\M\in\mathbb{M}$ is a permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{|\mathbb{M}|} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma^2).
  \label{eq:similarity}
\end{equation}

Additionally, we have the constraint of small
transformations. Specifically, if the current mental image is $\Xt$,
then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ rotate right by $\epsilon$,} \\
      f(\Xt, 0, -\epsilon) &\mbox{ rotate left by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip.} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}

To summarize, we approximate the likelihood term of
Eq.~\ref{eq:lh-ratio} using the similarity function defined in
Eq.~\ref{eq:similarity}. Because we assume analog rotations, this
similarity can only be computed at sequential locations, i.e. the
actions listed in Eq.~\ref{eq:actions}.

\subsection{Specific models}

The last decision we need to make is which algorithm to use to
approximate Eq.~\ref{eq:lh-ratio} using samples of the similarity
function. The algorithm must do two things: first, it must choose
\textit{which} places to sample, and second, it must decide when to
stop sampling.

\subsubsection{\Oc{} model}

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. For each $h$, we then check whether the resulting matrix
is a valid rotation matrix.

\subsubsection{\Th{} model}

An intuitive and simple model which does not know which way to rotate
\textit{a priori} might first rotate in the direction that increases
similarity, and then stop rotating once a ``match'' is found, as
determined by a threshold on the value of $S$. If all rotations have
been exhausted, then flip, and try rotating again.

We note, however, that it is not always clear what constitutes a good
enough ``match''. In the current formulation of the problem, we know
the exact geometry of the shapes, and know that a linear
transformation exists which will perfectly align the shapes. However,
shapes could be more complex and/or three-dimensional, be subject to
perceptual uncertainty, or even be not exactly identical. In these
cases, it is not clear how to choose a reasonable threshold \textit{a
  priori}, as the global optimum will depend on many factors. Thus, it
is unlikely that the \Th{} model will not be robust beyond this
domain.

The consequence of doing a search for a match is that
Eq.~\ref{eq:lh-ratio} becomes a likelihood ratio test, where $\theta$
is set to the MLE value, rather than being marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimisation
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. However, it is not necessarily the case that
all local maxima for the correct hypothesis will be higher than all
the local maxima for the incorrect hypothesis: thus, we would expect
the \Hc{} model to have difficulties reliably choosing the correct
hypothesis.

@@

Rather than a point estimate of the ratio, or even a simple
Monte Carlo estimate, we can infer a distribution by placing priors
over the similarity functions themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors, as in \citeA{Osborne:2012tm}
\TODO{probably need to describe better what a GP is}. Because our
domain is on a circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
$\log(S_h)$. Because GPs have no guarantee of non-negativity, placing
the prior over the log enforces positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP.\footnote{Note that the approach
  used here differs slightly from the one presented in
  \citeA{Osborne:2012tm} and is based on personal communication with
  one of the authors \cite{Duvenaud:2013td}.}We will denote these
points as $\bar{S}_h:=\exp(\mu_h)$. Then, from
\citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

Now that we have a distribution over $Z_h$, we can additionally
compute which points which we would expect to strengthen our estimate
(i.e., decrease the variance). From \citeA{Osborne:2012tm}, the
expected variance of $Z_h$ if we sampled a new observation at
$\theta_a$ is:
\begin{align}
\mathbb{E}&[\mathbb{V}(Z_h|\theta_a)]=\mathbb{V}(Z_h) + \mathbb{E}[Z_h] - \\
&\int \mathbb{E}[Z_h|\theta_{a}]^2 \mathcal{N}(\mu_h(\theta_a), \mathrm{Cov}_h(\theta_a, \theta_a))\dif\log S_h(\theta_a).\nonumber
\label{ref:expected-variance}
\end{align}

Revisiting our likelihood ratio, we now have:
\begin{equation}
p(\ell)=p\left(\frac{Z_0}{Z_1}\right)\approx\frac{\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0], \mathbb{V}(Z_0))}{\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1], \mathbb{V}(Z_1))}.
\end{equation}
This distribution does not have a nice form. However, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$: thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$. Assuming $Z_0$ and $Z_1$
are independent,\footnote{This assumption is probably not
  correct. \TODO{explanation?}} we have:
\begin{equation}
  p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1], \mathbb{V}(Z_0) + \mathbb{V}(Z_1)).
\end{equation}
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$h=0$, and when $p(Z_D<0)>0.975$, we accept $h=1$.

>>>>

\section{Modeling mental rotation}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $\hi$ is the hypothesis that the images
depict the same object, and $\hf$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta,
  \label{eq:posterior}
\end{equation}
where $p(X_b\ \vert\ X_a, \theta, h)$ is the probability that $X_b$
was generated from $X_a$, and $p(X_a)$ is the probability of observing
$X_a$. Because we ultimately want to determine which hypothesis is
more likely, the quantity of interest is a ratio $\ell:=p(\hi\ \vert\
X_a, X_b) / p(\hf\ \vert\ X_a, X_b)$ which (assuming that all
rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\left(\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta\right)p(\hi)}{\left(\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta\right)p(\hf)}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

We represent a shape of $N$ vertices with a $N\times 2$ coordinate
matrix $\mathbf{X}=[\mathbf{x}_1, \ldots{}, \mathbf{x}_N]$, and denote
the rotation and/or reflection transformation as $f(\mathbf{X}, h,
\theta):=\mathbf{X}\F_h^T\R^T$, where $\R$ is a rotation matrix, and
$\F_h$ is either the identity matrix $\I$ (when $\hi$) or a reflection
matrix across the $y$-axis (when $\hf$).

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$:
\begin{equation}
  p(\Xb\ \vert\ \Xa, \theta, h):= S(\Xb, f(\Xa, h, \theta)).
  \label{eq:likelihood}
\end{equation}
We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings,
$\mathbb{M}$. For brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h,
\theta)$ where $\M\in\mathbb{M}$ is a permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{|\mathbb{M}|} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $\sigma_S=0.15$ is the standard deviation of the similarity.

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, |\epsilon|) &\mbox{ rotate by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \mathcal{N}(0, \sigma_\epsilon)$ and
$\sigma_\epsilon=0.7$ is the standard deviation of the step size in
radians.

To summarize, we approximate the likelihood term of
Eq.~\ref{eq:lh-ratio} using the similarity function defined in
Eq.~\ref{eq:similarity}. Because we assume mental rotations are
performed sequentially, this similarity can only be computed for the
actions listed in Eq.~\ref{eq:actions}.

\subsection{Specific models of mental rotation}

In order to approximate Eq.~\ref{eq:lh-ratio} using samples of the
similarity function, we must decide \textit{which} places to sample
and \textit{when} stop sampling. The models below differ in how they
make these decisions.

\subsubsection{Oracle model}

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if computation produces a
valid rotation matrix; the $h$ that does produce a rotation matrix is
the correct hypothesis. This approach gives us the true value of
$\theta$, so Eq.~\ref{eq:lh-ratio} becomes a likelihood ratio test,
where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)p(\hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)p(\hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

\subsubsection{Threshold model}

An intuitive and simple model which does not know which way to rotate
\textit{a priori} might first rotate in the direction that increases
similarity, and then stop rotating once a ``match'' is found, as
determined by a threshold on the value of $S$. If all rotations have
been exhausted, then flip, and try rotating again. We assume the value
of $S$ which is greater than the threshold corresponds to the true
$\theta$, or a point near the true $\theta$. So, as with the \Oc{}
model, Eq.~\ref{eq:lh-ratio} becomes a likelihood ratio.

We note that in the current formulation of the problem, choosing the
threshold is straightforward because we know both the exact geometry
of the shapes and that a linear transformation exists which will
perfectly align them. However, it is not always clear how to choose a
reasonable threshold \textit{a priori}, as the global optimum will
depend on many factors (such as shape complexity, dimensionality,
perceptual uncertainty, and whether the shapes are exactly
identical). Thus, it is unlikely that the \Th{} model will be robust
beyond this domain.

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimisation
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. As with the \Oc{} and \Th{} models,
Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood ratio. Unlike the
global maximum, however, it is not guaranteed that a local maximum of
the correct hypothesis will be greater than a local maximum of the
incorrect hypothesis.

@@

Formally, we denote $Z_h$ as our estimate of the likelihood for
hypothesis $h$, and write its distribution as:
\begin{equation}
  p(Z_h) = \int \left[\int S(\Xb, f(\Xa, \theta, h))p(\theta)\dif\theta\right] p(S)\dif S.
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use a Gaussian Process (GP) prior for $p(S)$, as in
\citeA{Osborne:2012tm}.\footnote{Because our domain is on a circle, we
  utilize a periodic kernel for the covariance
  \cite{Rasmussen:2006vz}, i.e. $k(\theta,
  \theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.}
Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
the log of $S_h$ in order to enforce positivity after it is
exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

Assuming independence, we can now write $p(Z_h)\approx\mathcal{N}(Z_h\
\vert\ \mathbb{E}[Z_h], \mathbb{V}(Z_h))$, which gives us a
distribution over the likelihood ratio in Eq.~\ref{eq:lh-ratio}:
\begin{equation}
p(\ell)\approx\frac{\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0], \mathbb{V}(Z_0))\ p(\hi)}{\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1], \mathbb{V}(Z_1))\ p(\hf)}.
\end{equation}
This distribution cannot easily be calculated; however, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$. Thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$:
\begin{equation}
  p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1], \mathbb{V}(Z_0) + \mathbb{V}(Z_1)).
\end{equation}
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$\hi$, and when $p(Z_D<0)>0.975$, we accept $\hf$. To determine which
points to sample, we can compute the expected variance of $Z_h$ if we
sampled a new observation at $\theta_a$. From \citeA{Osborne:2012tm}:
\begin{align}
\mathbb{E}&[\mathbb{V}(Z_h|\theta_a)]=\mathbb{V}(Z_h) + \mathbb{E}[Z_h] - \\
&\int \mathbb{E}[Z_h|\theta_{a}]^2 \mathcal{N}(\mu_h(\theta_a), \mathrm{Cov}_h(\theta_a, \theta_a))\dif\log S_h(\theta_a).\nonumber
\label{ref:expected-variance}
\end{align}
We compute this expected variance for each of the actions in
Eq.~\ref{eq:actions}, and then choose the action with the lowest
value.
